Fail Faster

### 2025-03-20 paper/paper.tex
newswarm
fix
eval

Generators are traditionally handwritten, but some libraries allow users to synthesize 
them automatically from type definitions. Type-derived generators are highly 
convenient---the derivation process requires no manual effort---but also limitated: 
they are unable to account for constraints absent from the type. 
As a result, generating correct values with complex preconditions is difficult.
This trade-off can be deceptive, especially for inexperienced PBT users, who might derive a generator, run their test harness, find no bugs, and assume the code is correct---when in reality, most values are never tested at all because they do not satisfy the preconditions.

@@

%\jwc{NOTE: we should test generator speed across both languages, but speed -> bugfinding ability in only OCaml.}
%\jwc{
%  Baseline generators to test speed in both languages:
%  \begin{itemize}
%    \item Single int
%    \item Pair of ints, constrained
%    \item List of ints without bind (use map): lots of sampling, minimal binds.
%    \item Unableled Trees of a fixed size (no weighted union) minimal sampling, lots of binds.
%  \end{itemize}
%}
\subsection{Benchmarking Bug-Finding Ability}
To answer \textbf{RQ2}, we evaluate the bug-finding ability of our BST and STLC
case studies using Etna, a platform for assessing and comparing PBT techniques. 
Etna allows users to measure the effectiveness of different generator
implementations by injecting bugs into the system under test and recording the
time taken for a relevant property to fail in response. For example, in a tree
with an ordering property, a bug injection might consist of inserting a value
into the tree that violates that ordering. Each case study includes a diverse
set of \textit{tasks}, where a task consists of a specific bug-property pair
designed to test different aspects of the system (e.g., BST includes tasks that
test insertion, deletion, and union operations on binary search trees). In particular, BST has 43 tasks, and STLC has 18.

Strategies for BST and STLC are unchanged from \secref{subsection:benchmarking}.
To ensure a fair comparison, we initialize each strategy with the same random
seed, ensuring that they produce identical sequences of values. We run each
strategy on all tasks using a 60-second timeout.  If a bug is not found within
this limit, the task is considered unsolved.  Tasks where all strategies fail to
find the bug are excluded from our dataset, as they provide no basis for
comparison.  Similarly, we exclude tasks where the BQ strategy completes in
under 5ms, as such negligible runtimes do not yield meaningful insights into
relative performance. \tr{We probably want to discuss how many tasks get filtered out...} These filters ensure that reported speedups are based on
cases where optimization has a measurable impact.

It is worth noting that the time-to-failure of a given strategy on a given seed
is not necessarily representative of the strategy's average time-to-failure over
a large number of trials. We normalize by computing the relative performance, or
the ``speedup''. This works in most cases, but it is not perfect.  For example,
it is theoretically possible to choose a seed such that the \textit{first} value
produced by a generator discovers the bug, which would elide any differential
speedup (and likewise if both generators fail to find the bug).  To account for
these cases, we repeat the above process over 10 \tr{(soon to be 30)} seeds.\footnote{Although the
variance \textit{between} seeds in Etna can be vast, timing results are
replicable for a \textit{given} seed. Across 1,000 trials using the same seed, 
the observed variance in time-to-failure was less than a nanosecond.}

The average individual-task speedups for each strategy and benchmark are shown in 
\figref{fig:etna}, with trends in bug-finding ability closely mirroring those in performance 
from \secref{subsection:benchmarking}. The distribution of speedups in \figref{fig:swarm} reveals 
substantial variability, with most tasks clustering near the median and a long tail of outliers 
achieving much larger gains. STLC shows a more bimodal distribution of speedups
than the other benchmarks, which we attribute to the heterogenous difficulty of its tasks:  
some tasks regularly hit the 60-second timeout, while others finish in a fraction of a second. We find that tasks of intermediate difficulty 
tend to achieve larger speedups, whereas very fast or slow tasks cluster near the mean. \tr{To be honest, we don't know EXACTLY why STLC exhibits this bimodal distribution. Maybe it's better to leave this out, and just say ``look, there's high variance, but it's actually a good thing---sometimes you get more than you expect, and you almost never get less than you expect"?}
Overall, 
these results give strong evidence that our technique consistently benefits bug-finding ability, sometimes drastically.
\begin{figure}[h]
  \includegraphics[scale=0.333]{speedup.png}
  \caption{Geometric average of all speedups---relative to BQ---for each strategy and benchmark, showing that staging leads to better bug-finding ability across the board.}
  \label{fig:etna}
\end{figure}

\begin{figure}[h]
  \includegraphics[scale=0.26]{swarm.png}
  \caption{Swarm distribution of individual-task speedups across strategies and benchmarks, black dots represent tasks. Plots with extremely low variance have no swarm overlay; in plots with high variance, outliers correspond to large speedups. \tr{Make mean higher-contrast.}}
  \label{fig:swarm}
\end{figure}

>>>>

Generators are traditionally handwritten, but some libraries allow users to synthesize 
them automatically from type definitions. Type-derived generators are 
convenient---the derivation process requires no manual effort---but also limited: 
they are unable to account for constraints not encoded in the type. 
As a result, generating values with complex preconditions is statistically unlikely.
This state of affairs is not ideal, as inexperienced PBT users, who might derive a generator, run their test harness, find no bugs, and assume the code is correct---when in reality, most values are never tested at all because they do not satisfy the preconditions.

@@

%\jwc{NOTE: we should test generator speed across both languages, but speed -> bugfinding ability in only OCaml.}
%\jwc{
%  Baseline generators to test speed in both languages:
%  \begin{itemize}
%    \item Single int
%    \item Pair of ints, constrained
%    \item List of ints without bind (use map): lots of sampling, minimal binds.
%    \item Unableled Trees of a fixed size (no weighted union) minimal sampling, lots of binds.
%  \end{itemize}
%}
\subsection{Benchmarking Bug-Finding Ability}
To answer \textbf{RQ2}, we evaluate the bug-finding ability of our BST and STLC
case studies using Etna, a platform for assessing and comparing PBT techniques. 
Etna allows users to measure the effectiveness of different generator
implementations by injecting bugs into the system under test and recording the
time taken for a relevant property to fail in response. For example, in a tree
with an ordering property, a bug injection might consist of inserting a value
into the tree that violates that ordering. Each case study includes a diverse
set of \textit{tasks}, where a task consists of a specific bug-property pair
designed to test different aspects of the system (e.g., BST includes tasks that
test insertion, deletion, and union operations on binary search trees). 
In particular, BST has 37 tasks, and STLC has 20. Strategies for BST and STLC 
are implemented in
\camlname; they are unchanged from their description in \secref{subsection:benchmarking}.\footnote{Etna does 
not support Scala, so we were unable to test \scalaname, but we see no reason
we see no reason our results in this section would not extend to \scalaname.}

It is worth noting that the time-to-failure of a given strategy on a given seed
is not necessarily representative of the strategy's average time-to-failure over
a large number of trials. We normalize by computing the relative performance, or
the ``speedup''. This works in most cases, but it is not perfect.  For example,
it is theoretically possible to choose a seed such that the \textit{first} value
produced by a generator discovers the bug, which would elide any differential
speedup (and likewise if both generators fail to find the bug).  To account for
these cases, we repeat the above process over 30 random seeds---that is, all strategies
are run on the same seed so that they produce identical sequences of values, and this process
is repeated 30 times.\footnote{Although the
variance \textit{between} seeds in Etna can be vast, timing results are
replicable for a \textit{given} seed. Across 1,000 trials using the same seed, 
the observed variance in time-to-failure was less than a nanosecond.}

We run each
strategy on all tasks using a 60-second timeout.  If a bug is not found within
this limit, the task is considered unsolved.  Tasks where all strategies fail to
find the bug are excluded from our dataset, as they provide no basis for
comparison.  Similarly, we exclude tasks where the BQ strategy completes in
under 5ms, as such negligible runtimes do not yield meaningful insights into
relative performance. These filters remove
28/600 (4.67\%) of type-derived STLC's tasks, 
168/600 (28\%) of STLC's tasks, 
149/1110 (14.42\%) of type-derived BST's tasks,
268/1110 (24.1\%) of single-pass BST's tasks, and
371/1110 (33.42\%) of insertion-based BST's tasks. More sophisticated 
strategies tend to find bugs faster, leading to a higher number of filtered 
tasks. By applying these filters, we ensure that our reported speedups reflect 
optimizations that meaningfully impact performance.

The average individual-task speedups for each strategy and benchmark are shown in 
\figref{fig:etna}, with trends in bug-finding ability closely mirroring those in performance 
from \secref{subsection:benchmarking}. The distribution of speedups in \figref{fig:swarm} reveals 
substantial variability, with most tasks clustering near the median and a long tail of outliers 
achieving much larger gains. STLC shows a more bimodal distribution of speedups
than the other benchmarks, which we attribute to the heterogenous difficulty of its tasks:  
some tasks regularly hit the 60-second timeout, while others finish in a fraction of a second. We 
find that ``easier'' tasks---those that run on the order of milliseconds---regularly achieve speedups in the 
range of 300 to 800\%, whereas tasks that run on the order of seconds achieve more moderate speedups of up to 250\%.
Overall, 
these results give strong evidence that \camlname consistently benefits bug-finding ability, sometimes drastically.
\begin{figure}[h]
  \includegraphics[scale=0.333]{speedup.png}
  \caption{Geometric average of all speedups---relative to BQ---for each strategy and benchmark, showing that staging leads to better bug-finding ability across the board.}
  \label{fig:etna}
\end{figure}
\begin{figure}[h]
  \includegraphics[scale=0.35]{swarm.png}
  \caption{Swarm distribution of individual-task speedups across strategies and benchmarks, gray dots represent tasks. Plots with extremely low variance have no swarm overlay; in plots with high variance, outliers correspond to large speedups. \tr{Make mean higher-contrast.}}
  \label{fig:swarm}
\end{figure}
